---
title: "Exploring the Diamonds Dataset"
description: |
  Using the diamonds dataset to explore ggplot functions and customization more.
author:
  - name: Paloma Cartwright
    url: {}
date: 08-10-2021
output:
  distill::distill_article:
    self_contained: false
preview: images/diamonds.png
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(skimr)
library(GGally)
```
## Using GGally and ggplot to Explore the Diamonds data set 


### Using ggplot 

#### Creating a Scatter Plot 

Creating a scatter plot of carat versus price of the diamonds data set. 

```{r}

ggplot(data = diamonds, aes(x = carat, y = price)) +
  geom_point() +
  aes(color = color) +
  facet_wrap(~cut)
  
```
Generally, as the carats increase, the price of the diamonds also increase. 

#### Creating a Histogram 

Creating a histogram of the carats of diamonds in the dataset 

```{r}
ggplot(data = diamonds, aes(x = carat, fill = color)) +
  geom_histogram() +
  aes(color = carat)
```


### Using GGally 

This outputs a really ridiculous set of graphs, but that is the purpose of the ggpairs function. 

```{r}
GGally::ggpairs(diamonds)
```



