---
title: "Map Making"
description: |
  This is a blog post that is my journey on how to make maps.
author:
  - name: Paloma Cartwright
    url: {}
date: 09-29-2021
categories: 
  - R
  - spData
  - tmap
output:
  distill::distill_article:
    toc: true
    code_folding: true
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(sf)
library(raster)
library(dplyr)
library(spData)
library(spDataLarge)
library(tmap)   
library(leaflet) 
library(tidyverse)
library(RColorBrewer)
library(rgeos)
library(rnaturalearth)
library(rnaturalearthdata)

```

## Let's Make Some Maps 
### tmap basics

This is a blog post about my process navigating chapter 8 in GeoComputation with R for EDS 223. I don't know where this journey will take me, but I figured that I'd make a blog post about my process as I make some cool maps and learn how to use tmap. 


Making a basic map of New Zealand with tmap. The New Zealand object comes from the sf data 

`tm_polygons()` combines both `tm_fill()` and `tm_borders()` so map_nz1 and map_nz2 below are the same.

```{r}
map_nz1 <- tm_shape(nz) +
  tm_fill() + 
  tm_borders() 

map_nz2 <- tm_shape(nz) + tm_polygons()
map_nz2
```

Combining multiple layers 

`tm_raster()` plots a raster layer and the argument alpha is the transparency of that layer 

```{r}
map_nz3 <- map_nz2 + tm_shape(nz_elev) +
  tm_raster(alpha = 0.7) 
map_nz3
```

New Zealand water addition 

```{r}
nz_water <- st_union(nz) %>% #returns a single geometry with resolved boundaries
  st_buffer(22200) %>% #computes a buffer around this geometry
  st_cast(to = "LINESTRING") #cast geometry to another type

map_nz4 <- map_nz3 + 
  tm_shape(nz_water) + tm_lines()
map_nz4


```

Adding the height element to the map of New Zealand 

```{r}
map_nz5 <- map_nz4 + 
  tm_shape(nz_height) + tm_dots()

map_nz5
```

To combine different maps using `tmap` in a similar way as you would use `patchwork` in ggplot you do the following: 

```{r}
tmap_arrange(map_nz3, map_nz4, map_nz5)
```

### tmap aesthetics 

The common aesthetics for fill and border layers are color `col` , transparency `alpha` , line width `lwd` and line type`lty`.

```{r}
map1 <- tm_shape(nz) + tm_fill(col = "green")
map2 <- tm_shape(nz) + tm_fill(col = "red", alpha = 0.5)
map3 <- tm_shape(nz) + tm_borders(col = "darkgrey")
map4 <- tm_shape(nz) + tm_borders(lwd = 2)
map5 <- tm_shape(nz) + tm_borders(lty = 3)
map6 <- tm_shape(nz) + tm_fill(col = "green", alpha = 0.5) +
  tm_borders(col = "darkgrey", lwd = 2, lty = 3)
tmap_arrange(map1, map2, map3, map4, map5, map6)
```

You cannot use a column name in tmap like you can in R to set the aesthetic of a variable like color, but instead you have to put it in quotations. You can treat plot the same way as you would R though. Hopefully, I am understanding this correctly. 

```{r}
# tm_shape(nz) + tm_fill(col = nz$Land_area) This line fails
plot(st_geometry(nz), col = nz$Land_area) # This works because plot has the same functionality as R
tm_shape(nz) + tm_fill(col = "Land_area") #The color aesthetic has to be in quotes here 

```

To add a title in the legend of the aesthetic you are using, you define this in the same place as you define the aesthetic. 
`expression()` is used in the commented seciton below to add the special script for the squared kilometers. If you do not need fancy text, you can just put your title in quotes. 

```{r}
#legend_title = expression("Area (km"^2*")") 
legend_title = "Area in Kilometers Squared"
map_nza = tm_shape(nz) +
  tm_fill(col = "Land_area", title = legend_title) + tm_borders(lwd = 1.5)
map_nza
```

### Color Settings 

```{r}
tm_shape(nz) + tm_polygons(col = "Median_income")
breaks = c(0, 3, 4, 5) * 10000
tm_shape(nz) + tm_polygons(col = "Median_income", breaks = breaks)
tm_shape(nz) + tm_polygons(col = "Median_income", n = 10)
tm_shape(nz) + tm_polygons(col = "Median_income", palette = "viridis")
```

This is all the map work I've done for the day! Stay tuned, maybe there will be more. 

