---
title: "Hurricanes and GDP Growth in The Bahamas"
description: |
  EDS 222 Final Project: Exploring the statistical relationship between Hurricanes and the Annual Percentage of   GDP Growth in The Bahamas.
author:
  - name: Paloma Cartwright
    url: https://palomacartwright.github.io/
date: 2021-11-24
categories: 
  - R
  - statistics
  - tidyverse
output:
  distill::distill_article:
    toc: true
    code_folding: true
    self_contained: false
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(here)
library(janitor)
library(patchwork)
library(kableExtra)
library(gt)
library(xtable)
library(broom)
```


# Research Question

The aim of this project is to investigate the impact of hurricanes on the annual percentage of GDP growth in The Bahamas. Growing up in The Bahamas, I have witnessed first hand the devastating impacts of hurricanes on livelihoods throughout the country and have been an active participant in recovery efforts every time my island was struck. Natural disasters have massive economic and human costs and take a deep toll on growth prospects because reconstruction costs monopolize already scarce resources for health, education, and social spending, especially in developing countries (Ötker & Srinivasan, 2018). 

There is existing research on the impact of hurricanes on GDP growth in the United States but this project will look specifically at The Bahamas. With the global changing climate and the continual impact of stronger hurricanes on Caribbean countries, answering this question can have important implications for the future of small island developing states like The Bahamas. 


# Data Collection and Tidying 

For the analysis of this research question, data from the National Hurricane Center and The World Bank are being used. 

## GDP Growth Data 

The data for percentage of GDP Growth was collected from The World Bank Data Repository. This data covers annual percentage of GDP growth from 1961 to present for The Bahamas. The metadata containing information about calculations and sampling strategy can be found [here](https://databank.worldbank.org/reports.aspx?source=2&type=metadata&series=NY.GDP.MKTP.KD.ZG).

- World Bank. (2021). GDP growth (annual %)—Bahamas, The | Data. https://data.worldbank.org/indicator/NY.GDP.MKTP.KD.ZG?locations=BS

### Data I/O and Tidying

```{r}
gdp_growth <- read_csv(here("_posts", "2021-11-24-hurricanes-and-gdp-growth-in-the-bahamas", "final_proj_data", "bah_yearly_gdp_growth.csv")) %>% 
  rename("country_name" = "Country Name") %>% 
  filter(country_name == "Bahamas, The") %>% 
  select(-"Indicator Name") %>% 
  pivot_longer(cols = 2:62, 
               names_to = "year", 
               values_to = "gdp_growth") %>% 
  mutate(year = as.numeric(year))
```

### Data Visualization

```{r}
ggplot(data = gdp_growth, aes(x = year, y = gdp_growth)) +
  geom_line(color = "lightblue") +
  geom_point(color = "blue") +
  labs(x = "Year", 
       y = "Annual % GDP Growth", 
       title = "Yearly Percentage of GDP Growth for The Bahamas") +
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```
## Hurricane Data

The data on hurricanes in The Bahamas was collected from the National Hurricane Center's Data Archive (National Hurricane Center, 2021). This dataset did not exist in csv format, so I looked through the yearly reports from 1960 to present and recorded all of the storms that impacted The Bahamas. I made note of the year each storm occurred, the name of the storm, the intensity and the number of islands in The Bahamas that were impacted by each storm. Because The Bahamas is an archipelago of 17 major islands, I wanted to note how many were impacted by each storm because that would be a metric that could possibly affect GDP growth. If there are multiple islands that need to be rebuilt, this would have a greater impact.

In creation of this dataset, I recorded storm intensity from 1 to 6, with 1 being Tropical Storms and 6 being a category 5 hurricane. The Saffir-Simpson Wind Scale considers a tropical storm anything with winds below 74mph (National Hurricane Center, 2021), but I know that these storm can still have damaging impacts on countries so I wanted to ensure they were included in the data to be analyzed for this question. 

### Data I/O, Tidying and Visualization

```{r}
storms <- read_csv(here("_posts", "2021-11-24-hurricanes-and-gdp-growth-in-the-bahamas", "final_proj_data", "bahamas_hurricanes.csv")) 

yr_storms <- storms %>% 
  group_by(year) %>% 
  summarize(number = n(), 
            avg_strength = mean(category),
            avg_numisland = mean(no_of_islands))

yr_storms_summary <- yr_storms %>% 
  kable(
    digits = 2,
    col.names = c("Year", "Number of Storms", "Average Strength", "Average Number of Islands"), 
    align = 'lccr',
    caption = "Bahamian Hurricane Summary Statistics 1960 - 2021")

yearly <- ggplot(data = yr_storms, aes(x = year, y = number)) +
  geom_point(color = "blue") + 
  labs(x = "Year", 
       y = "Number of Storms", 
       title = "Yearly Number of Storms to Impact the Bahamas") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

intensity <- ggplot(data = yr_storms, aes(x = year, y = avg_strength)) +
  geom_point(color = "blue") +
  labs(x = "Year", 
       y = "Average Intensity of Storms", 
       title = "Average Yearly Storm Intensity") +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

yr_storms_summary

yearly / intensity

```

# Combining the Datasets 

To complete my analysis, I had to combine the GDP growth dataset with the Hurricane dataset. 

```{r}
hgdp <- full_join(gdp_growth, yr_storms, by = 'year')
```

# Analysis Plan 

Clearly describe your analysis plan. What is your analysis plan? Why did you choose this analysis, given your data and question? What are the limitations?

For this analysis I'm going to first look at a linear regression of hurricanes on percentage of GDP growth. I will then do hypothesis testing on the regression and to explore further. 

```{r}
ggplot(data = hgdp, aes(x = number, y = gdp_growth)) +
  geom_jitter() + 
  geom_smooth(method = 'lm', se = F)
```


```{r}
lm(gdp_growth ~ number*avg_strength, data = hgdp) %>%
  summary() %>% 
  xtable() %>% 
  kable()

mod <- lm(gdp_growth ~ number + avg_strength, data = hgdp)
```

```{r}
hgdp %>% 
  ggplot(aes(y = gdp_growth, x = number, color = avg_strength)) +
  geom_point() +
  stat_smooth(method="lm", se=FALSE) + 
  labs(x = "Average Number of Storms",
       y = "Rate of GDP Growth")
```



# Delete Later 



- World Bank. (2021). GDP (current US$)—Bahamas, The | Data. https://data.worldbank.org/indicator/NY.GDP.MKTP.CD?locations=BS

```{r}
gdp <- read_csv(here("_posts", "2021-11-24-hurricanes-and-gdp-growth-in-the-bahamas", "final_proj_data", "Bah_yearly_gdp.csv")) %>% 
  rename("country_name" = "Country Name") %>% 
  filter(country_name == "Bahamas, The") %>% 
  select(-"Indicator Name") %>% 
  pivot_longer(cols = 2:62, 
               names_to = "year", 
               values_to = "gdp") %>% 
  mutate(year = as.numeric(year))

complete_gdp <- full_join(gdp, gdp_growth, by = c("country_name", "year"))

```


